Chasing \rainbows{} with the Habitable Worlds Observatory

### 2023-04-06 src/tex/ms.tex
[showyourwork] overleaf sync

@@

% Authors are in alphabetical order:
\author[Sophia R. Vaughan et al.]{%
    Sophia R. Vaughan$^{1,}$\thanks{Correspondence:  \url{sophia.vaughan@physics.ox.ac.uk}}
    Kimberly Bott$^{2,28}$,
    Sarah L. Casewell$^{3}$,
    Nicolas B. Cowan$^{4}$,
    David S. Doelman$^{5}$,\newauthor 
    Timothy D. Gebhard$^{6}$,
    Matthew Kenworthy$^{5}$,
    Johan Mazoyer$^{7}$,
    Maxwell A. Millar-Blanchaer$^{8}$,\newauthor 
    Olivier Absil$^{16}$,
    Lisa Altinier$^{13}$,
    Pierre Baudoz$^{7}$,
    Ruslan Belikov$^{24}$,
    Alexis Bidot$^{9}$,\newauthor 
    Markus Johannes Bonse$^{6}$,
    Bernhard Brandl$^{5}$,
    Alexis Carlotti$^{9}$,
    Elodie Choquet$^{13}$,
    Niyati Desai$^{17}$,\newauthor
    Kevin Fogarty$^{24}$,
    Jules Fowler$^{12}$
    Yann Gutierrez$^{7,14,22}$,
    Olivier Guyon$^{15,25,26,27}$,
    Sebastiaan Y. Haffert,$^{15}$, \newauthor
    Olivier Herscovici-Schiller$^{14}$,
    Adrien Hours$^{9}$,
    Roser Juanola-Parramon $^{20,21}$,
    Elina Kleisioti $^{5}$, \newauthor
    Lorenzo König$^{16}$,
    Mariya Krasteva$^{23}$,
    Iva Laginja$^{7}$,
    Rico Landman$^{5}$,
    Lucie Leboulleux$^{9}$,
    David Mouillet$^{9}$,\newauthor
    Mamadou N’Diaye$^{10}$,
    Emiel H. Por$^{18}$,
    Laurent Pueyo$^{18}$,
    Frans Snik$^{5}$,
    Dirk van Dam$^{5}$,
    Kyle van Gorkom, $^{15}$,\newauthor
    Maaike van Kooten$^{11}$
    \newauthor 
    \\
    $^{1}$ Department of Astrophysics, University of Oxford, Denys Wilkinson Building, Keble Road, Oxford OX1 3RH, UK\\
    $^{2}$ Department of Earth and Planetary Sciences, University of California, Riverside, CA 92521, USA \\
    $^{3}$ Centre for Exoplanet Research, School of Physics and Astronomy, University of Leicester, University Road, Leicester, LE1 7RH, UK\\
    $^{4}$ Department of Earth \& Planetary Sciences and Department of Physics, McGill University, 3600 rue University, Montréal, QC, H3A 2T8, Canada\\
    $^{5}$ Leiden Observatory, Leiden University, P.O. Box 9513, 2300 RA Leiden, The Netherlands\\
    $^{6}$ Max Planck Institute for Intelligent Systems, Max Planck Ring 4, 72076 Tübingen, Germany \& ETH Zurich, Switzerland \\
    $^{7}$ LESIA, Observatoire de Paris, Universit\'e PSL, CNRS, Sorbonne Universit\'e, Universit\'e de Paris, F-92195 Meudon, France \\
    $^{8}$ Department of Physics, University of California, Santa Barbara, CA 93106, USA \\
$^{9}$ Univ. Grenoble Alpes, CNRS, IPAG, 38000 Grenoble, France \\
$^{10}$ Universit\'e C\^ote d'Azur, Observatoire de la C\^ote d'Azur, CNRS, Laboratoire Lagrange. France \\
$^{11}$ National Research Council Canada, Herzberg Astronomy and Astrophysics Research Center, Victoria, B.C. Canada \\
$^{12}$ University of California, Santa Cruz  \\
$^{13}$ Aix Marseille Univ, CNRS, CNES, LAM, Marseille, France \\
$^{14}$ DTIS, ONERA, Université Paris Saclay, 91123 Palaiseau, France \\
$^{15}$ Steward Observatory, University of Arizona, 933 North Cherry Avenue, Tucson, Arizona \\
$^{16}$ STAR Institute, Universit\'e de Li\`ege, All\'ee du Six Ao\^ut 19c, 4000 Li\`ege, Belgium \\
$^{17}$ Department of Astronomy, California Institute of Technology, Pasadena, CA, 91125, USA \\
$^{18}$ Space Telescope Science Institute, 3700 San Martin Drive, Baltimore, MD 21218, USA \\
$^{19}$ SRON Netherlands Institute for Space Research, Niels Bohrweg 4, 2333 CA, Leiden, The Netherlands \\
$^{20}$ NASA Goddard Space Flight Center, 8800 Greenbelt Rd, Greenbelt, MD  20771,United States \\
$^{21}$ University of Maryland Baltimore County, 1000 Hilltop Cir, Baltimore, MD 21250, United States \\
$^{22}$ DOTA, ONERA, 92322 Ch\^atillon, France \\
$^{23}$ European Space Agency, ESTEC, Keplerlaan 1, 2200 AG Noordwijk, The Netherlands \\
$^{24}$ NASA Ames Research Center, Bldg. 245, Moffett Field, USA \\
$^{25}$ Subaru Telescope, NAOJ, United States \\
$^{26}$ College of Optical Sciences, University of Arizona, Tucson, AZ 87521, United States \\
$^{27}$ Astrobiology Center, 2 Chome-21-1, Osawa, Mitaka, Tokyo, 181-8588, Japan \\
$^{28}$ NASA Nexus for Exoplanet System Science, Virtual Planetary Laboratory Team, Seattle, WA 98195, USA}
\date{Draft version: \today}

@@

\begin{abstract}
NASA recently announced the Habitable Worlds Observatory, a coronagraphic mission which would detect rocky planets in the habitable zone, establish their habitability, and search them for biosignatures. 
%Numerous instrumental and observational challenges have still to be overcome to design a coronagraphic mission with that goal. After a research phase aiming at the detection of a few tens of exo-Earth in the habitable zone, the Habitable Worlds Observatory will use its spectroscopic and polarimetric capabilities to probe these planets' atmosphere and surface properties. 
Surface liquid water is central to the definition of planetary habitability.
%
Photometric and polarimetric phase variations can show the presence of specularly reflecting oceans on exoplanets, in addition to being sensitive to rainbows and other cloud scattering effects. Direct imaging missions are optimised to detect planets near quadrature, which may obscure some of the phases at which these features are strongest. The range of scattering phases accessible for an exoplanet can be limited by its orbital inclination or the coronagraph's inner working angle. Planets that orbit close to edge-on have accessible phase angles limited by obscuration by the coronagraph. We use the list of target stars for the Habitable Worlds Observatory to estimate the number of exo-Earths that could be searched for non-Lambertian scattering phenomena. We find that exo-Earths in systems listed in this Habitable Worlds Observatory catalog will have a detectable Rayleigh scattering peak. The glint signature at scattering phases of $\sim$50--70$^\circ$ would be accessible to the Habitable Worlds Observatory in $\sim$20 systems, and the rainbow of water clouds at phases of 140--160$^\circ$ would be accessible in $\sim$55 systems, assuming a 60 mas inner working angle (\IWA). The number of systems in which an exo-Earth could be searched for these scattering phenomena increases with orbital eccentricity and scales approximately inversely with \IWA.

\end{abstract}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec:intro}

\begin{figure*}%[t]
   \centering
   \includegraphics[width=\linewidth]{figures/Fig1IWAPol.png}
   \caption{BOTT PLOT this is a holder until Kim updates the figure}
    \label{fig:bottplot}
\end{figure*}

@@

Surface liquid water is closely related to planetary habitability: water is required for life as we know it, and life on exoplanets will most likely on be detectable if water is present at the surface. 
To first order, the presence of surface liquid water can be predicted by the flux a planet receives from its star. The habitable zone (HZ) describes the range of semi-major axes around a given main sequence star for which an Earth-like planet could maintain surface liquid water \citep{kasting93}.  The HZ is convenient because a planet can be deemed habitable based on solely on its orbital separation, but it remains an untested hypothesis. We would therefore like to establish the presence of liquid water on the surface of an exoplanet before trying to interpret its atmospheric biosignatures. %Ideally, we will empirically establish the presence of liquid water on an exoplanet before trying to interpret its putative biosignatures.

@@

\subsection{Ocean Glint, Cloud Glory and Rainbows}
\cite{2001Natur.412..885F} noted the potential for time-resolved reflected light photometry to identify water oceans on the surface of an exoplanet.  \cite{2008Icar..195..927W} performed simulations of scattered light from directly imaged terrestrial exoplanets. 
%to demonstrate that phase-resolved photometry and polarization of a directly-imaged exoplanet would betray the presence of large surface oceans. 
They showed that oceans would be detectable because they are more reflective and exhibit a peak in polarization at crescent phases. %Rotational variability of an exoplanet's reflected light colours can be inverted into a rough map of the planet's surface, with dark regions being suggestive of oceans \citep{2009ApJ...700..915C}, but
Specular reflection, a.k.a. glint, remains the most direct means to infer liquids on the surface of other worlds \citep[e.g., Titan's methane lakes;][]{2010GeoRL..37.7104S}. %\citet{lustig2019} explored mapping exoplanets at crescent phases using ocean glint.
Specular reflection appears at similar phase angles in both unpolarized and polarized light and its peak orbital phase can be extended by atmospheric winds, which produce waves and hence increase the size of the glint spot \citep{2011Icar..211..722B,kopparla2018}.
%\cite{2006astro.ph.10518M} quantified how photometric and polarimetric scattered light orbital phase variations might betray the presence of oceans on terrestrial exoplanets.  
\citet{Takahashi2021} used polarimetric Earthshine measurements to detect the presence of the ocean, proving that the method works in practice for an unresolved world.

%\noindent {\color{Green}<Modifications by Victor Trees>}

%\cite{2008Icar..195..927W} simulated photometric and polarimetric phase curves of directly imaged terrestrial planets. They noted that surface oceans would be detectable because they are more reflective and exhibit a peak in polarization at crescent phases.

% NEEDS Paragraph on mapping connection to tectonics

% NEEDS KOPPARLA (prob refer to , BERDYUGINA, ROSSI? (cloud?), HU?, added... blend cloud hinderance into argument for context to lead into rainbows, mention Lietzow and jeremy there., maybe some more mention of the Earthshine observations Karalidi, Gordon, etc

%"Stam & Hovenier (2006) independently examined the observability of the polarized signatures of an Earth-like extrasolar planet, including Rayleigh scattering and sea-surface glint. Schneider (2006) had similar ideas." -KB

% Move to Clouds?
In practice, clouds complicate the detection of ocean glint because they obscure the underlying surface and can act as strong forward scatterers, mimicking ocean glint in the total  reflected flux (polarized plus unpolarized). \cite{2010ApJ...721L..67R} compared observations of the Earth's scattering phase curve to radiative transfer simulations; they found that the photometric brightening of Earth at crescent phases is mostly due to forward scattering from clouds, with a relatively small contribution from ocean glint.

\cite{Zugger_2010} simulated photometric and polarized phase variations at visible wavelengths for Earth-like planets including the effects of %partial cloud coverage
Rayleigh scattering and a horizontally homogeneous cloud cover. They concluded that detection of an ocean glint signature in polarization is severely limited by the presence of clouds. %Although the presence of clouds and atmospheric Rayleigh scattering weaken the signal of an underlying ocean, they %again 
%find that oceans are detectable and distinguishable from other sources of polarization. 
%{\color{Green} 
\citet{treesstam2019} simulated the total and polarized fluxes and the degree of polarization of light reflected by ocean planets with patchy clouds. They found that the alternation of cloud-free and cloud-covered glint, as the planet rotates and its cloud cover changes in time, can result in strong fluctuations in the degree of polarization, hence resulting in noisier phase curves. Nonetheless, polarization phase curves of ocean planets at optical and near infrared wavelengths intersect \citep{treesstam2019}, providing a possible avenue to distinguish between an exo-ocean and forward scattering clouds.%}%}

%{\color{Green}
The Rayleigh scattering optical thickness of a planet's atmosphere is smaller in the near-infrared (NIR) than in the visible, which improves the visibility of the glint \citep{Zugger_2011}. NIR phase measurements also minimize the latitude--albedo effect, which could otherwise be a false-positive for photometric brightening at crescent phases due to glint \citep{2012ApJ...752L...3C}.

Phase variations of polarized light could also reveal other non-Lambertian cloud scattering phenomena. The degree of polarization of light scattered by spherical cloud droplets strongly depends on the scattering angle \citep[see][]{hansentravis1974}, and consequently, on the phase angle for an Earth-like exoplanet with liquid water clouds \citep[see][]{stam2008}. The sign of the degree of polarization (i.e. the direction of polarization) changes sign, and show peaks at the primary and secondary rainbow angles  \citep{bailey2007,2018MNRAS.480.1613B}. A measurement of those rainbows would be a strong indicator of spherically shaped particles in the exoplanet's atmosphere. The precise phase angle locations in the the degree of polarization of the rainbow peaks depend on wavelength \citep[see Fig. 5 of][]{trees2022}, the particle size ...  and on the type of liquid \citep{bailey2007}. A famous example... Venus ... %A famous example of using polarimetry to characetrize clouds within our Solar System is the characterization of the main cloud deck of Venus.  \citep{hansenhovenier1974} derived that the main cloud deck of Venus consists of a 75\% sulfuric acid solution.
%
... glory ...\citep{2014A&A...566L...1G} ... and halos ... ice clouds ...

 \textbf{Victor/Kim: please complete this paragraph with a few sentences describing what one learns from rainbow and glory, ideally backed with good references!}

@@

%\subsection{Rainbows}
%calc; retrieve species including for non terrestrials; droplet liklihood of liquid surface; lit review bailey et below

%The rai

%cloud coverage paper in polarisation :
%\cite{rossi17} computed flux and polarisation of reflected starlight for different types of cloud cover on Earth-like planets and determined different types of cloud cover can be distinguished from each other. %Alternative options or papers here.... Kim can write this tomorrow , needs other Stam, Karalidi, Kopparla, maybe Gordon, need to reference rainbows in Bailey paper.  Look at citations in Gordon paper (our bib)... he has a really thorough lit review --KB

@@

%\textbf{In this paper we...}
\subsection{Outline}
In this paper we quantify the accessible scattering phase angles for hypothetical terrestrial planets orbiting in the habitable zones of HabWorlds target stars. In \S 2.1 we describe the target list, in \S 2.2 we derive expressions for the range of scattering phase angles in a few different limits, in \S 2.3 we describe Monte Carlo simulations to marginalize over the unknown orbital inclination and eccentricity of planets in these systems, and in \S 2.4 we outline how the planet/star contrast changes as a function of orbital phase.  We present and discuss our results in \S 3 and conclude in \S 4.  %the   the way that inclination and coronagraph inner working angle can limit the accessible phase angles  
% We minimally explore the contrast ratio as we are primarily exploring whether these scattering angles are reachable with inner working angle and outer working angle considered.

@@

\begin{figure}
    \centering
    \includestandalone{tikz/annotated-orbit}
    \caption{
        The coronagraph can obscure the exoplanet at high and low scattering phase angles. This can prevent observations of scattering phenomena which could be used to indicate the habitability and atmospheric composition of the exoplanet. \todo{Add why scattering is cool} 
    }
    \label{fig:annotated-orbit}
\end{figure}

@@

\Cref{fig:scatterplot} shows $\Delta \varphi$ for the systems in the HabWorlds target list assuming an IWA of 62 mas ($3 \lambda / D$ coronagraph at $\SI{600}{\nano\meter}$). 
%
%This assumes each planet has a semi-major axis such that it receives an Earth-equivalent instellation, which is provided in the input star list.
%
For edge-on circular orbits, it will be possible to observe Rayleigh scattering for all targets and rainbows will be accessible in some cases.

>>>>

% Patch \hypertarget; for details, see 
% https://tex.stackexchange.com/a/17138/97669
\makeatletter
\newcommand{\affiliationtarget}[1]{\Hy@raisedlink{\hypertarget{#1}{}}}
\makeatother
\newcommand{\affiliationlink}[1]{\hyperlink{#1}{#1}}

@@

% Authors are in alphabetical order:
\author[Sophia R. Vaughan et al.]{%
    Sophia R. Vaughan$^{\affiliationlink{1},}$\thanks{Correspondence:  \url{sophia.vaughan@physics.ox.ac.uk}}
    Kimberly Bott$^{\affiliationlink{2},\affiliationlink{28}}$,
    Sarah L. Casewell$^{\affiliationlink{3}}$,
    Nicolas B. Cowan$^{\affiliationlink{4}}$,
    David S. Doelman$^{\affiliationlink{5,19}}$,
    \newauthor 
    Timothy D. Gebhard$^{\affiliationlink{6},\affiliationlink{29}}$,
    Matthew Kenworthy$^{\affiliationlink{5}}$,
    Johan Mazoyer$^{\affiliationlink{7}}$,
    Maxwell A. Millar-Blanchaer$^{\affiliationlink{8}}$,
    \newauthor 
    Olivier Absil$^{16}$,
    Lisa Altinier$^{13}$,
    Pierre Baudoz$^{7}$,
    Ruslan Belikov$^{24}$,
    Alexis Bidot$^{9}$,
    Markus J. Bonse$^{29}$,
    \newauthor 
    Bernhard Brandl$^{5}$,
    Alexis Carlotti$^{9}$,
    Elodie Choquet$^{13}$,
    Niyati Desai$^{17}$,
    Kevin Fogarty$^{24}$,
    Jules Fowler$^{12}$,
    \newauthor
    Yann Gutierrez$^{7,14,22}$,
    Olivier Guyon$^{15,25,26,27}$,
    Sebastiaan Y. Haffert$^{15}$,
    Olivier Herscovici-Schiller$^{14}$, 
    \newauthor
    Adrien Hours$^{9}$,
    Roser Juanola-Parramon$^{20,21}$,
    Elina Kleisioti $^{5}$,
    Lorenzo König$^{16}$,
    Mariya Krasteva$^{23}$, 
    \newauthor
    Iva Laginja$^{7}$,
    Rico Landman$^{5}$,
    Lucie Leboulleux$^{9}$,
    David Mouillet$^{9}$,
    Mamadou N’Diaye$^{10}$,
    Emiel H. Por$^{18}$,
    \newauthor
    Laurent Pueyo$^{18}$,
    Frans Snik$^{5}$,
    Dirk van Dam$^{5}$,
    Kyle van Gorkom$^{15}$,
    Maaike van Kooten$^{11}$ 
    \newauthor \\%
    Affiliations are listed at the end of the paper
    % Timmy: I checked MNRAS and it seems that this is how they handle very long author lists. See, for example:
    % https://academic.oup.com/mnras/article/498/2/2354/5900562
    % I've also added some LaTeX magic to create links for the affiliations; if y'all agree that it's a good idea I can extend it to all authors?
}
\date{Draft version: \today}

@@

\begin{abstract}
NASA recently announced the Habitable Worlds Observatory (HWO), a coronagraphic mission to detect rocky planets in habitable zones, establish their habitability, and search for biosignatures. 
%Numerous instrumental and observational challenges have still to be overcome to design a coronagraphic mission with that goal. After a research phase aiming at the detection of a few tens of exo-Earth in the habitable zone, the Habitable Worlds Observatory will use its spectroscopic and polarimetric capabilities to probe these planets' atmosphere and surface properties. 
Surface liquid water is central to the definition of planetary habitability.
Photometric and polarimetric phase variations can show the presence of specularly reflecting oceans on exoplanets, in addition to being sensitive to rainbows and other cloud scattering effects. 
Direct imaging missions are optimised to detect planets near quadrature, which may obscure some of the phases at which these features are strongest. 
The range of scattering phases accessible for an exoplanet can be limited by its orbital inclination or the coronagraph's inner working angle. 
Planets that orbit close to edge-on have accessible phase angles limited by obscuration by the coronagraph. 
We use the list of target stars for the Habitable Worlds Observatory to estimate the number of exo-Earths that could be searched for non-Lambertian scattering phenomena. 
We find that exo-Earths in systems listed in this Habitable Worlds Observatory catalog will have a detectable Rayleigh scattering peak. 
The glint signature at scattering phases of $\sim$50--70$^\circ$ would be accessible to the Habitable Worlds Observatory in $\sim$20 systems, and the rainbow of water clouds at phases of 140--160$^\circ$ would be accessible in $\sim$55 systems, assuming a 60\,mas inner working angle (\IWA). 
The number of systems in which an exo-Earth could be searched for these scattering phenomena increases with orbital eccentricity and scales approximately inversely with \IWA.     \end{abstract}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec:intro}

%--------------------------------------------------------------------------------
\begin{figure*}%[t]
   \centering
   \includegraphics[width=\linewidth]{figures/Fig1IWAPol.png}
   \caption{BOTT PLOT this is a holder until Kim updates the figure {\color{Green}Victor Trees: The glory (in the backscattering direction) and diffraction peak (in the forward scattering direction) are distinct features in the total reflected flux, but not in the degree of polarization. In principle, the degree of polarization goes to zero in the forward and backward scattering directions. See e.g. Fig. 1 of \citet{stam2008}, Fig. 5 of \citet{trees2022}, Fig. 3 of \citet{karalidi2011} and \citet{hansentravis1974}.}}
    \label{fig:bottplot}
\end{figure*}
%--------------------------------------------------------------------------------

@@

Liquid water is closely related to planetary habitability because it is essential for life as we know it. 
Indeed, to be able to detect such life forms, this liquid water should be on the surface of an exoplanet instead of below it.
To first order, the presence of surface liquid water can be predicted by the flux a planet receives from its star. The habitable zone (HZ) describes the range of semi-major axes around a given main sequence star for which a rocky planet would have the right temperature to maintain liquid surface water \citep{kasting93}. 
The HZ is a convenient concept because with it a planet can be deemed habitable based on solely its orbital separation, but it remains an untested hypothesis. 
We therefore prefer to establish the presence of liquid surface water on an exoplanet before attempting to interpret biosignatures in the planet's atmosphere. 
% Ideally, we will empirically establish the presence of liquid water on an exoplanet before trying to interpret its putative biosignatures.

@@

\subsection{Ocean Glint, Cloud Glory and Rainbows}
%
Starlight incident on liquid water on the surface of an Earth-like exoplanet is reflected back into space towards the observer. This Fresnel (or specular) reflection leaves a mirror image of the star on the water surface, called the glint, which broadens with increasing water surface roughness driven by the wind speed \citep{CoxMunk1954}. The strength of the Fresnel reflection increases with increasing reflection angle locally on the planet, resulting in a stronger glint. Consequently, in the total reflected flux by the exoplanet as a whole, the signal of the glint is most prominent at the largest planetary phase angles.
%Within our Solar System, the glint has been used to detect methane lakes on Titan} \citep[][]{2010GeoRL..37.7104S}.

\cite{2001Natur.412..885F} noted the potential for time-resolved reflected 
light photometry to identify landmasses and oceans of liquid on the surface of 
a rotating exoplanet. \cite{2010ApJ...721L..67R} compared observations of the Earth's scattering phase curve to radiative transfer simulations; they found that the photometric brightening of Earth at crescent phases is mostly due to clouds, with a relatively small contribution from ocean glint. In practice, clouds complicate the detection of ocean glint because they can obscure the underlying surface and can act as strong forward scatterers, mimicking the ocean glint in the total reflected flux at large phase angles. Scattering of light by the atmospheric gas, i.e. Rayleigh scattering, also blurs the glint. The Rayleigh optical thickness of a planet's atmosphere is smaller in the near-infrared (NIR) than in the visible, which improves the visibility of the glint \citep{Zugger_2011}. NIR phase measurements also minimize the latitude--albedo effect, which could otherwise be a false-positive for photometric brightening at crescent phases due to glint \citep{2012ApJ...752L...3C}.

 %Rotational mapping of an exoplanet at gibbous phases would reveal oceans as dark regions \citep{2009ApJ...700..915C} whose albedo would increase sharply at crescent phases \citep{lustig2019}.

Except for only the total reflected flux at various wavelengths, the degree of polarization (the polarized flux divided by the total flux) and direction of polarization of the light could also be measured. The degree of polarization has practical advantages, because it is a relative measure: instrument errors could be cancelled out upon division of the polarized flux by the total flux, depending on the telescope design. As the starlight of a solar-type star is often considered to be unpolarized \citep{kemp1987}, and gets polarized during the reflection by the exoplanet, measuring polarized light increases the contrast between the exoplanet and its host star. The degree of polarization itself also provides information about the exoplanet's atmosphere and surface. In fact, most scattering phenomena in the atmosphere are more sensitive to the scattering angle in the degree of polarization than in the total (polarized plus unpolarized) flux \citep{hansentravis1974}. Also for detecting exo-oceans, measuring the degree of polarization could be an asset. The glint is a strong source of polarized reflected flux, and the degree of polarization of the light reflected by the (cloud-free) water surface peaks at the Brewster angle, which occurs at larger angular planet-star separations than the glint in the total flux \citep[see e.g.][]{Zugger_2010,treesandstam2019}.

%to demonstrate that phase-resolved photometry and polarization of a directly-imaged exoplanet would betray the presence of large surface oceans. 
%They showed that oceans would be detectable because they are more reflective and exhibit a peak in polarization at crescent phases (Figure~\ref{fig:bottplot}). 
%Specular reflection appears at similar phase angles in both unpolarized and polarized light and its peak orbital phase can be extended by atmospheric winds, which fuel surface waves and hence broaden the size of the glint spot 
%\cite{2006astro.ph.10518M} quantified how photometric and polarimetric scattered light orbital phase variations might betray the presence of oceans on terrestrial exoplanets.  
%\cite{2008Icar..195..927W} simulated photometric and polarimetric phase curves of directly imaged terrestrial planets. They noted that surface oceans would be detectable because they are more reflective and exhibit a peak in polarization at crescent phases. 
% NEEDS Paragraph on mapping connection to tectonics
% NEEDS KOPPARLA (prob refer to , BERDYUGINA, ROSSI? (cloud?), HU?, added... blend cloud hinderance into argument for context to lead into rainbows, mention Lietzow and jeremy there., maybe some more mention of the Earthshine observations Karalidi, Gordon, etc
%"Stam & Hovenier (2006) independently examined the observability of the polarized signatures of an Earth-like extrasolar planet, including Rayleigh scattering and sea-surface glint. Schneider (2006) had similar ideas." -KB

\cite{2008Icar..195..927W} performed simulations of reflected light from directly imaged exoplanets with oceans, and found strong peaks in the degree of polarization at the Brewster angles, but did not take into account Rayleigh scattering gas on top of the oceans. \cite{Zugger_2010} simulated photometric and polarized phase variations at visible wavelengths for ocean planets including the effects of Rayleigh scattering gas and a horizontally homogeneous cloud cover. They concluded that detection of an ocean glint signature in polarization is severely limited by the presence of clouds. \citet{kopparla2018} explored the ocean glint signatures in the signal from a multi-planetary (TRAPPIST-1) system, with TRAPPIST-1e being the ocean planet. They concluded that, although measuring in polarized light would increase the contrast between the ocean planet and the total signal of the multi-planetary system, the glint detectability is limited by Rayleigh scattering and clouds on the ocean planet and its neighbour planets.

%Although the presence of clouds and atmospheric Rayleigh scattering weaken the signal of an underlying ocean, they %again 
%find that oceans are detectable and distinguishable from other sources of polarization. 
%{\color{Green} 
\citet{treesandstam2019} simulated the total and polarized fluxes and the degree of polarization of light reflected by ocean planets with a \textit{patchy} cloud cover. They showed that the alternation of cloud-free and cloud-covered glint, as the cloud pattern changes in time, can result in strong fluctuations in the degree of polarization phase curves at continuum wavelengths (i.e. at wavelengths at which the light is not absorbed by gases in the exoplanet's atmosphere), with generally high polarization 
when the glint is visible and low polarization when the glint is covered. Fluctuations of the polarization continuum phase curves are also expected when continents rotate over the glint \citep{groot2020}. If the telescope's integration time is too long to resolve those temporal variations, the fluctuations will results in noisier phase curves. Nonetheless, the polarization phase curves of partially cloudy planets with oceans at optical and near infrared wavelengths intersect \citep{treesandstam2019}, providing a possible avenue to distinguish between an exo-ocean and forward scattering clouds.

\textit{<Observing the Earth as an exoplanet>}
Measurements of Earthshine (i.e. the reflected sunlight by the Earth incident on the dark side of the Moon) indeed reported higher degrees of polarization in the continuum at long wavelengths than at short wavelengths, which could be attributed to the ocean glint \citep{Emde2017,sterzik2019,takahashi2021}.

%\textit{\citet{sterzik2019} and \citet{Takahashi2021} used polarimetric Earthshine 
%measurements to detect the presence of the ocean, proving that the method 
%works in practice for a spatially unresolved world.}

The optical properties of an Earth-like exoplanet orbiting its host star may vary along the phase curve, e.g., because of the planet's seasons and changing weather systems in time. Recent simulations by \citet{trees2022} showed that signatures of the glint could also be found in the degree of polarization spectra (i.e., as functions of wavelength with 1 nm resolution), measured at a single phase angle. Taking into account the absorption of light by gases in the exoplanet's atmosphere, they computed that the degree of polarization of ocean planets is lower across gaseous absorption bands with respect to the continuum, while for dry planets, it is higher \citep[see][]{stam2008}. Those dips, instead of peaks, can already be searched for at intermediate phase angles ($90^\circ$-$130^\circ$), thus at the largest angular planet-star separations, and also occur for high planetary cloud fractions as long as the glint is (partially) cloud-free. Comparing the degree of polarization in the continuum and at a wavelength (band) where gases partly absorb the light, could potentially allow for an ocean detection with only one measurement in time. In this paper, we neglect absorption by gases and do not focus on phase variations of the spectra, but on the phase variations at single wavelengths in the continuum.

%Phase variations of polarized light could also reveal other non-Lambertian scattering phenomena (Figure~\ref{fig:bottplot}).

The degree of polarization of light scattered by spherical cloud droplets strongly depends on the scattering angle \citep[see][]{hansentravis1974}, and consequently, on the phase angle for an Earth-like exoplanet with liquid water clouds \citep{stam2008}. The scattering angle dependent variations of the degree of polarization for different droplet sizes are more distinct than in the total reflected flux \citep{karalidi2011}, and the sign of the degree of polarization (i.e. the direction of polarization) changes at various angles \citep{stam2008,trees2022}. In particular, the degree of polarization shows peaks at the primary and secondary rainbow angles, caused by reflection of light inside the spherical droplets, and the shapes of the peaks depend on the wavelength, particle size (distribution) and type of liquid  \citep{hansentravis1974,bailey2007,bailey2018}. 
A measurement of rainbow features in polarized light would be a strong indicator of spherically shaped (i.e. liquid) particles in the exoplanet's atmosphere, as they are not expected to originate from reflection by the planet's surface or non-spherical (ice) cloud particles \citep{Karalidi2012ice}. In the total reflected flux, thus without the separation of the polarized light, the rainbow features are virtually unobservable \citep{karalidi2011}. The benefits of those rainbow features in polarized light for cloud characterization on other planets than the Earth have been proven by \citet{hansenhovenier1974}, who fitted model calculations through measurements of the degree of polarization phase curves of Venus. %Because Venus is an inner planet as seen from the Earth, it can be observed from the Earth at phase angles ranging from almost $0^\circ$ to $180^\circ$ . 
%They retrieved that the main cloud deck of Venus consists of a 75\% sulfuric acid solution, with a droplet effective radius of 1.05 $\mu$m and effective variance of 0.07.

%... glory ...\citep{2014A&A...566L...1G} ... and halos ... ice clouds ...

@@

\subsection{Rainbows {\color{Green} [Victor: The rainbows are already discussed in previous section.]}}

A well-characterized rainbow feature in the phase curve of a planet provides a wealth of information about the atmosphere and condensates. The rainbow feature is visible in both polarized and unpolarized phase curves, but is particularly prominent in polarized light because it is strongly polarized and the fractional polarization from other sources like Rayleigh scattering is reduced at narrow scattering angles where many rainbows occur. The location of the rainbow peak in the phase curve shifts depending on the refractive index of the condensate and thus will shift depending on the species, and for most species the peak will also shift slightly with wavelength. The relative height of the rainbow compared to the Rayleigh peak with wavelength provides evidence of the depth of the cloud top height. The size distribution of droplets also effects the polarized rainbow and other optical effects.
% Shape/phase
% 
% cite Bailey

As a characterization tool the rainbow and other cloud optical features are vital, and in studies of habitability they can provide evidence that water is the dominant condensate providing context to the glint feature.  In this work we identify our ideal rainbow angle as %...

In our own Solar System the additional information inherent in condensate polarization allowed the identification of the species and droplet size distribution of the clouds on Venus \citep{Hansen1974}, the discernment between a snowy surface and hazy thick atmosphere on Titan \citep{Zellner1973}, and may have eliminated some prospects for the unknown UV absorber on Venus \citep{Petrova2018}.

%\cite{rossi17} computed flux and polarisation of reflected starlight for different types of cloud cover on Earth-like planets and determined different types of cloud cover can be distinguished from each other. %Alternative options or papers here.... Kim can write this tomorrow , needs other Stam, Karalidi, Kopparla, maybe Gordon, need to reference rainbows in Bailey paper.  Look at citations in Gordon paper (our bib)... he has a really thorough lit review --KB

@@

%\textbf{In this paper we...}
\subsection{Outline}

In this paper, we quantify the phase angles that are accessible for direct
observations for hypothetical terrestrial planets orbiting in the habitable 
zones of HabWorlds target stars. 
In \S 2.1, we describe the target list, in \S 2.2 we derive expressions for the 
range of phase angles given a few different limits, in \S 2.3, we describe 
Monte Carlo simulations to marginalize over the unknown orbital inclination and 
eccentricity of planets in these systems, and in \S 2.4, we outline how the 
planet/star contrast changes as a function of the planet's orbital phase.  
We present and discuss our results in \S 3 and conclude in \S 4. 
%the   the way that inclination and coronagraph inner working angle can limit the accessible phase angles  
% We minimally explore the contrast ratio as we are primarily exploring whether these scattering angles are reachable with inner working angle and outer working angle considered.

@@

\begin{figure}
    \centering
    \includestandalone{tikz/annotated-orbit}
    \caption{
        An illustration of an edge-on circular orbit showing that the coronagraph can obscure the exoplanet at high and low scattering phase angles. This can prevent observations of scattering phenomena which could be used to indicate the habitability and atmospheric composition of the exoplanet. 
        %\timmy{Maybe mention that this is a simplified illustration with a circular orbit, plus we are basically looking at an edge-on system?}
    }
    \label{fig:annotated-orbit}
\end{figure}

@@

\Cref{fig:scatterplot} shows $\Delta \varphi$ for the systems in the \hwo\ target list, assuming an IWA of \SI{62}{\mas} ($3 \lambda / D$ coronagraph at $\SI{600}{\nano\meter}$). 
\timmy{60 (as in the abstract), or 62?}
%
%This assumes each planet has a semi-major axis such that it receives an Earth-equivalent instellation, which is provided in the input star list.
%
For edge-on circular orbits, it will be possible to observe Rayleigh scattering for all targets and rainbows will be accessible in some cases.
